---
title: "Practical Machine Learning Project"
author: "Liselle Caronongan"
date: "December 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(RGtk2)
```
## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity. These types of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are simply interested and techie. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.

In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Libraries
```{r}
library(caret)
library(lattice)
library(ggplot2)
library(rattle)
```
## Data load
```{r}
TrainData <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"),header=TRUE)
dim(TrainData)

TestData <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"),header=TRUE)
dim(TestData)

str(TrainData)
```

The training data set is made of 19622 observations on 160 columns. We see that many columns have NA values or blank values on almost every observation. Thus, we will remove them because they will not produce any substantial information. The first seven columns give information about the people who did the test and also timestamps. We will not take them in our model.

We get the indices of the columns having at least 90% of NA or blank values on the training dataset
```{r}
indColToRemove <- which(colSums(is.na(TrainData) |TrainData=="")>0.9*dim(TrainData)[1]) 
TrainDataClean <- TrainData[,-indColToRemove]
TrainDataClean <- TrainDataClean[,-c(1:7)]
dim(TrainDataClean)
```

We do the same for the test set
```{r}
indColToRemove <- which(colSums(is.na(TestData) |TestData=="")>0.9*dim(TestData)[1]) 
TestDataClean <- TestData[,-indColToRemove]
TestDataClean <- TestDataClean[,-1]
dim(TestDataClean)

str(TestDataClean)
```
After cleaning, the new training data set has only 53 columns.

We now create a partition of the traning data set 
```{r}
set.seed(12345)
inTrain1 <- createDataPartition(TrainDataClean$classe, p=0.75, list=FALSE)
Train1 <- TrainDataClean[inTrain1,]
Test1 <- TrainDataClean[-inTrain1,]
dim(Train1)

dim(Test1)
```
In the following sections, we will test 3 different models:
1. classification tree 
2. random forest 
3. gradient boosting method

In order to limit the effects of overfitting and improve the efficicency of the models, we will use the cross-validation technique. We will use 5 folds (usually, 5 or 10 can be used, but 10 folds gives higher run times with no significant increase of the accuracy).


```{r}
trControl <- trainControl(method="cv", number=5)
model_CT <- train(classe~., data=Train1, method="rpart", trControl=trControl)

print(model_CT)
fancyRpartPlot(model_CT$finalModel)
trainpred <- predict(model_CT,newdata=Test1)

confMatCT <- confusionMatrix(Test1$classe,trainpred)

# display confusion matrix and model accuracy
confMatCT$table

confMatCT$overall[1]
```
We see that the accuracy of the first model is very low (about 55%). This means that the outcome class will not be predicted very well by the other predictors.


## Train with random forests
```{r}
model_RF <- train(classe~., data=Train1, method="rf", trControl=trControl, verbose=FALSE)

print(model_RF)


plot(model_RF,main="Accuracy of Random forest model by number of predictors")


trainpred <- predict(model_RF,newdata=Test1)

confMatRF <- confusionMatrix(Test1$classe,trainpred)

# display confusion matrix and model accuracy
confMatRF$table
confMatRF$overall[1]
names(model_RF$finalModel)
model_RF$finalModel$classes

plot(model_RF$finalModel,main="Model error of Random forest model by number of trees")

# Compute the variable importance 
MostImpVars <- varImp(model_RF)
MostImpVars
```

With random forest, we reach an accuracy of 99.3% using cross-validation with 5 steps. This is already good. But let's see what we can expect with Gradient boosting.

We can also see that the optimal number of predictors, i.e. the number of predictors giving the highest accuracy, is 27. There is no significant increase of the accuracy with 2 predictors and 27, but the slope decreases more with more than 27 predictors (even if the accuracy is still very good). The fact that not all the accuracy is worse with all the available predictors allows us to suggest that there may be some dependencies between them.

Lastly, using more than about 30 trees does not reduce the error significantly.

##  Train with gradient boosting method
```{r}
model_GBM <- train(classe~., data=Train1, method="gbm", trControl=trControl, verbose=FALSE)
print(model_GBM)
plot(model_GBM)
trainpred <- predict(model_GBM,newdata=Test1)

confMatGBM <- confusionMatrix(Test1$classe,trainpred)
confMatGBM$table
confMatGBM$overall[1]
```
Precision with 5 folds is 95.9%.

## Conclusion
This shows that the random forest model is the best one. We will then use it to predict the values of classe for the test data set.
```{r}
FinalTestPred <- predict(model_RF,newdata=TestDataClean)
FinalTestPred
```

